---
title: "Project 1 Report Figures"
author: "Rachel Austin"
date: "2026-03-02"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{threeparttable}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tibble)
library(stringr)
library(knitr)
library(kableExtra)
library(purrr)
```

```{r, include = FALSE}
data_dir <- "/Users/austinra/BIOS6624/Project 1/Data Processed"

# Frequentist models
vload.mod        <- readRDS(file.path(data_dir, "frequentist.vload.mod.rds"))
vload.mod.adh    <- readRDS(file.path(data_dir, "frequentist.vload.mod.adh.rds"))

leu3n.mod        <- readRDS(file.path(data_dir, "frequentist.leu3n.mod.rds"))
leu3n.mod.adh    <- readRDS(file.path(data_dir, "frequentist.leu3n.mod.adh.rds"))

agg_ment.mod     <- readRDS(file.path(data_dir, "frequentist.aggment.mod.rds"))
agg_ment.mod.adh <- readRDS(file.path(data_dir, "frequentist.aggment.mod.adh.rds"))

agg_phys.mod     <- readRDS(file.path(data_dir, "frequentist.aggphys.mod.rds"))
agg_phys.mod.adh <- readRDS(file.path(data_dir, "frequentist.aggphys.mod.adh.rds"))

# Bayesian summary tables
summary_table_vload      <- readRDS(file.path(data_dir, "summary_table_vload.rds"))
summary_table_vload_adh  <- readRDS(file.path(data_dir, "summary_table_vload_adh.rds"))

summary_table_leu3n      <- readRDS(file.path(data_dir, "summary_table_leu3n.rds"))
summary_table_leu3n_adh  <- readRDS(file.path(data_dir, "summary_table_leu3n_adh.rds"))

summary_table_agg_ment     <- readRDS(file.path(data_dir, "summary_table_agg_ment.rds"))
summary_table_agg_ment_adh <- readRDS(file.path(data_dir, "summary_table_agg_ment_adh.rds"))

summary_table_agg_phys     <- readRDS(file.path(data_dir, "summary_table_agg_phys.rds"))
summary_table_agg_phys_adh <- readRDS(file.path(data_dir, "summary_table_agg_phys_adh.rds"))
```


Extracting frequentist summary table values
```{r, include = FALSE}
extract_freq <- function(mod, term_pattern = "^hard_drugs", conf_level = 0.95) {

  sm <- summary(mod)$coefficients
  rn <- rownames(sm)

  term <- rn[grepl(term_pattern, rn)]

  term <- term[1]

  est <- sm[term, "Estimate"]
  p   <- sm[term, "Pr(>|t|)"]

  ci <- confint(mod, level = conf_level)
  lo <- ci[term, 1]
  hi <- ci[term, 2]

  tibble(
    term = term,
    freq_est = as.numeric(est),
    freq_lo  = as.numeric(lo),
    freq_hi  = as.numeric(hi),
    freq_p   = as.numeric(p)
  )
}

fmt_ci <- function(lo, hi, digits = 3) sprintf("[%.*f, %.*f]", digits, lo, digits, hi)
fmt_p  <- function(p) ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
```

Extracting bayes summary table values
```{r, include = FALSE}
extract_bayes_beta3 <- function(sumtab) {
  df <- as.data.frame(sumtab)
  row <- df[df$Parameter == "beta[3]", , drop = FALSE]
  if (nrow(row) != 1) stop("Couldn't uniquely find beta[3] in Bayesian summary table.")

  tibble(
    bayes_est = as.numeric(row$Estimate),
    bayes_lo  = as.numeric(row$HPDI_2.5),
    bayes_hi  = as.numeric(row$HPDI_97.5)
  )
}
```
Creating table
```{r, include = FALSE}

make_row <- function(outcome_label, freq_mod, bayes_sumtab) {
  f <- extract_freq(freq_mod, term = "hard_drugs1")
  b <- extract_bayes_beta3(bayes_sumtab)

  tibble(
    Outcome = outcome_label,

    `Bayesian Difference` = b$bayes_est,
    `95% HPDI` = fmt_ci(b$bayes_lo, b$bayes_hi, digits = 3),

    `Frequentist Difference` = f$freq_est,
    `95% CI` = fmt_ci(f$freq_lo, f$freq_hi, digits = 3),
    `p-value` = fmt_p(f$freq_p)
  )
}

table_primary <- bind_rows(
  make_row("Log(Viral Load)",         vload.mod,    summary_table_vload),
  make_row("CD4+ T-Cell Count",       leu3n.mod,    summary_table_leu3n),
  make_row("Agg. Mental QOL Score",   agg_ment.mod, summary_table_agg_ment),
  make_row("Agg. Physical QOL Score", agg_phys.mod, summary_table_agg_phys)
)

table_primary

```

\setcounter{table}{1}
Formatting table
```{r results='asis'}

tab_pub <- table_primary %>%
  transmute(
    Outcome = Outcome,
    `Posterior mean` = `Bayesian Difference`,
    `95% HPDI` = `95% HPDI`,
    Estimate = `Frequentist Difference`,
    `95% CI` = `95% CI`,
    `p-value` = `p-value`
  )

tab_pub <- tab_pub %>%
  mutate(
    `Posterior mean` = round(`Posterior mean`, 3),
    Estimate = round(Estimate, 3)
  )

tab_pub %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    align = "lccccc",
    caption = "Model Results for Bayesian and Frequentist Analyses"
  ) %>%
  add_header_above(c(" " = 1, "Bayesian Model" = 2, "Frequentist Model" = 3)) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, width = "2.2in") %>%
  footnote(
general = "Estimates represent adjusted mean differences at 2-year follow-up (hard drug users − non-users), adjusted for baseline outcome and covariates. Non-users are the reference group for baseline hard drug use. Viral load was modeled on the log 10 scale. Bayesian estimates are posterior means with 95% HPDIs; frequentist estimates are linear regression coefficients with 95% confidence intervals.",
    threeparttable = TRUE
  )
```


Grabbing adherence model data
```{r,include = FALSE}
table_adh <- bind_rows(
  make_row("Log(Viral Load)",         vload.mod.adh,    summary_table_vload_adh),
  make_row("CD4+ T-Cell Count",       leu3n.mod.adh,    summary_table_leu3n_adh),
  make_row("Agg. Mental QOL Score",   agg_ment.mod.adh, summary_table_agg_ment_adh),
  make_row("Agg. Physical QOL Score", agg_phys.mod.adh, summary_table_agg_phys_adh)
)
```

Creating adherence model table

```{r results='asis'}
tab_pub_adh <- table_adh %>%
  transmute(
    Outcome = Outcome,
    `Posterior mean` = `Bayesian Difference`,
    `95% HPDI` = `95% HPDI`,
    Estimate = `Frequentist Difference`,
    `95% CI` = `95% CI`,
    `p-value` = `p-value`
  ) %>%
  mutate(
    `Posterior mean` = round(`Posterior mean`, 3),
    Estimate = round(Estimate, 3)
  ) %>%
  # avoid LaTeX unicode minus error if any
  mutate(across(where(is.character), ~ gsub("\u2212", "-", .x)))

tab_pub_adh %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    align = "lccccc",
    caption = "Model Results for Bayesian and Frequentist Analyses (Models Adjusting for Adherence)"
  ) %>%
  add_header_above(c(" " = 1, "Bayesian Model" = 2, "Frequentist Model" = 3)) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, width = "2.2in") %>%
  footnote(
    general = "Estimates represent adjusted mean differences at 2-year follow-up (hard drug users minus non-users), adjusted for baseline outcome, covariates, and adherence. Non-users are the reference group for baseline hard drug use. Viral load was modeled on the log10 scale. Bayesian estimates are posterior means with 95% HPDIs; frequentist estimates are linear regression coefficients with 95% confidence intervals.",
    threeparttable = TRUE
  )
```

Reading in posterior probability data
```{r, include = FALSE}
files <- c(
  "bayes_results_vload_primary.rds",
  "bayes_results_leu3n_primary.rds",
  "bayes_results_agg_ment_primary.rds",
  "bayes_results_agg_phys_primary.rds"
)

results_list <- map(file.path(data_dir, files), readRDS)

files_adh <- c(
  "bayes_results_vload_adh.rds",
  "bayes_results_leu3n_adh.rds",
  "bayes_results_agg_ment_adh.rds",
  "bayes_results_agg_phys_adh.rds")

results_list_adh <- map(file.path(data_dir, files_adh), readRDS)
```

Creating table
```{r, include = FALSE}
#chatgpt function to create table
extract_pclin_looic <- function(x) {
  tibble(
    outcome = x$outcome,
    model = x$model,
    p_clin = as.numeric(x$p_clin),
    looic = as.numeric(x$loo$estimates["looic", "Estimate"]),
    looic_se = as.numeric(x$loo$estimates["looic", "SE"])
  )
}

df_primary <- map_dfr(results_list, extract_pclin_looic)
df_adh     <- map_dfr(results_list_adh, extract_pclin_looic)


outcome_labels <- c(
  VLOAD = "Log(Viral Load)",
  LEU3N = "CD4+ T-Cell Count",
  AGG_MENT = "Agg. Mental QOL Score",
  AGG_PHYS = "Agg. Physical QOL Score"
)

thresholds <- tibble(
  outcome = c("VLOAD", "LEU3N", "AGG_MENT", "AGG_PHYS"),
  `Clinically meaningful (±Δ)` = c("±0.5", "±50", "±2", "±2")
)

table_clin_loo <- thresholds %>%
  mutate(Outcome = outcome_labels[outcome]) %>%
  left_join(df_primary %>% select(outcome, p_clin, looic), by = "outcome") %>%
  rename(
    `Primary: Posterior P(|β|>Δ)` = p_clin,
    `Primary: LOOIC` = looic
  ) %>%
  left_join(df_adh %>% select(outcome, p_clin, looic), by = "outcome") %>%
  rename(
    `+Adherence: Posterior P(|β|>Δ)` = p_clin,
    `+Adherence: LOOIC` = looic
  ) %>%
  select(
    Outcome,
    `Clinically meaningful (±Δ)`,
    `Primary: Posterior P(|β|>Δ)`, `Primary: LOOIC`,
    `+Adherence: Posterior P(|β|>Δ)`, `+Adherence: LOOIC`
  ) %>%
  mutate(
    `Primary: Posterior P(|β|>Δ)` = round(`Primary: Posterior P(|β|>Δ)`, 4),
    `+Adherence: Posterior P(|β|>Δ)` = round(`+Adherence: Posterior P(|β|>Δ)`, 4),
    `Primary: LOOIC` = round(`Primary: LOOIC`, 1),
    `+Adherence: LOOIC` = round(`+Adherence: LOOIC`, 1)
  )

table_clin_loo
```

```{r results='asis'}

table_clin_loo %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    align = "lcccccc",
    caption = "Posterior Probability of Clinically Meaningful Effects and LOO-CV by Outcome"
  ) %>%
  add_header_above(c(" " = 2, "Primary model" = 2, "Primary model + adherence" = 2)) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    full_width = FALSE,
    position = "center"
  ) %>%
  footnote(
    general = "Posterior probabilities are computed as $P(|\\beta| > \\Delta)$ using posterior draws. LOOIC is the leave-one-out information criterion (lower indicates better out-of-sample predictive performance).",
    threeparttable = TRUE,
    escape = FALSE
  )

