---
title: "Simulation"
author: "Rachel Austin"
date: "2026-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(future)
library(future.apply)
library(furrr)
library(hdrm)
library(glmnet)
library(dplyr)
```
Potential Cases:
For all 
$\beta_1$ = 0.5/3
$\beta_2$ = 1/3
$\beta_3$ = 1.5/3
$\beta_4$ = 2/3
$\beta_5$ = 2.5/3
$\beta_6 - \beta_20$ = 0

N = 250 and N = 500
  a. $\rho$ = 0
  b. $\rho$ = 0.35
  c. $\rho$ = 0.7
  
Methods:
1. Backwards elimination using $\alpha$ = 0.05
2. Backwards elimination using AIC
3. Backwards elimination using BIC
4. LASSO
  a. $\lambda$.min
  b. $\lambda$.1se
5. Elastic Net


Functions Needed
```{r message = F, warning = F}
# Results extraction function for step function (p-values, AIC and BIC)
harvest <- function(mod_obj      # result model object from the reduction
                   ,param        # simulation target betas
                   ,alpha = 0.05 # crit for coverage and error computation
) {
  
  #extracting coefficients
  estimates <- as.data.frame(summary(mod_obj)$coefficients) 
  #new column called param stores estimate names (previously the rownames)
  estimates <- within(estimates, {
    param <- row.names(estimates)
    })
  #Wald CIs
  CIs <- as.data.frame(confint.default(mod_obj, level = 1-alpha))
  #change name from numeric vals
  names(CIs) <- c('LCL','UCL')
  #new column called param stores estimates names (previously rownames)
  CIs <- within(CIs, {
    param <- row.names(CIs)
  })
  
  #create varname objects
  var_names <- sprintf("V%02d", 1:length(param))
  
  # assemble summaries
  res_dat <- data.frame(
    #varnames of simulated dat
    variables = var_names
    #target betas we assigned in beginning are dropped into true_values
   ,true_values = param
  ) |>
    within({
      # flag whether v01-v20 shows up in estimates, only variables that were not dropped in selection step will show up here
      selected <- ifelse(variables %in% row.names(estimates), 1, 0)
      # flag whether variables v01-v20 are statistically significant 
      signif   <- ifelse(variables %in% row.names(estimates[estimates[,4] < alpha,]), 1, 0)
      # flag the first five var as true non-zero coefs
      true_non_zero <- ifelse(variables %in% sprintf("V%02d", 1:5), 1, 0)
    }) |>
    merge(CIs
          ,by.x = 'variables'
          ,by.y = 'param'
          ,all.x = T) |>
    within({
      #if the LCL is greater than the target betas assigned and the UCL is less than the target betas assigned we have coverage
      covered <- ifelse(LCL <= true_values & true_values <= UCL, 1, 0)
    })
    
  #if covered is missing, variable was not selected and variable is a true nonzero effect then covered = 0
  # aka model failed to select a variable that matters 
  res_dat[is.na(res_dat$covered) == T & 
          res_dat$selected == 0 & 
          res_dat$true_non_zero == 1,'covered'] <- 0 
  #if covered is missing, variable was not selected and variable is a true zero effect then covered = 1
  #aka model correctly excluded a null variable
  res_dat[is.na(res_dat$covered) == T & 
          res_dat$selected == 0 & 
          res_dat$true_non_zero == 0,'covered'] <- 1 
  
  return(res_dat)
}

#Results extraction function for glmnet function (LASSO and elastic net)
harvest_glmnet <- function(cv_obj,
                           param,
                           method,
                           n,
                           rho,
                           s = "lambda.min") {
  
  # extract coefficients at chosen lambda
  coef_obj <- coef(cv_obj, s = s)
  
  coef_df <- data.frame(
    variables = rownames(coef_obj),
    estimate  = as.numeric(coef_obj)
  )
  
  # remove intercept
  coef_df <- subset(coef_df, variables != "(Intercept)")
  
  # build results table
  res_dat <- data.frame(
    variables   = sprintf("V%02d", 1:length(param)),
    true_values = param
  )
  
  # align estimates with variable list
  res_dat$estimate <- coef_df$estimate[
    match(res_dat$variables, coef_df$variables)
  ]
  
  # unselected variables have coefficient 0
  res_dat$estimate[is.na(res_dat$estimate)] <- 0
  
  # same indicators as harvest()
  res_dat$true_non_zero <- ifelse(
    res_dat$variables %in% sprintf("V%02d", 1:5),
    1,
    0
  )
  
  res_dat$selected <- ifelse(res_dat$estimate != 0, 1, 0)
  
  # glmnet does not give p-values or Wald CIs
  res_dat$signif  <- NA
  res_dat$LCL     <- NA
  res_dat$UCL     <- NA
  res_dat$covered <- NA
  
  # metadata
  res_dat$method <- method
  res_dat$n      <- n
  res_dat$rho    <- rho
  
  # reorder columns to match your p-value output
  res_dat <- res_dat[, c(
    "variables",
    "true_values",
    "true_non_zero",
    "signif",
    "selected",
    "LCL",
    "UCL",
    "covered",
    "method",
    "n",
    "rho"
  )]
  
  return(res_dat)
}

```



## Simu function: generating data and running analyses

In this function I propose to encode the process of running one simulation iteration collecting and returning the results.

```{r}

simfunc <- function(n, rho) {
  
  # create simulated data
  param <- c(1/6, 1/3, 1/2, 2/3, 5/6, rep(0, 15))
  dat_obj <- gen_data(n = n 
                 ,p = 20
                 ,p1 = 5
                 ,beta = param
                 ,family = 'gaussian'
                 ,corr = 'exchangeable'
                 ,rho = rho
                 ) 
  dat <- data.frame(y=dat_obj$y, dat_obj$X)

  # reduction by p-values, AIC, BIC
  sat_model <- lm(y ~ ., data = dat)

  ## p-values
  pval_redux <- step(sat_model
                    ,direction = 'backward'
                    ,trace = 0
                    ,k = qchisq(1-0.05, 1)
                    )
  pval_res <- harvest(mod_obj = pval_redux  # result model object from the reduction
                     ,param   = param       # simulation target betas
                     ,alpha   = 0.05        # crit for coverage and error computation
  )
  pval_res$method <- 'PVAL'
  pval_res$n      <- n
  pval_res$rho    <- rho
  
  ## AIC
  AIC_redux <- step(sat_model
                    ,direction = 'backward'
                    ,trace = 0
                    ,k = 2
                    )
  AIC_res <- harvest(mod_obj = AIC_redux  # result model object from the reduction
                     ,param   = param       # simulation target betas
                     ,alpha   = 0.05        # crit for coverage and error computation
  )
  AIC_res$method <- 'AIC'
  AIC_res$n      <- n
  AIC_res$rho    <- rho

  ## BIC
  BIC_redux <- step(sat_model
                    , direction = 'backward'
                    , trace = 0
                    , k = log(nrow(dat)))
  
  BIC_res <- harvest(mod_obj = BIC_redux
                     , param = param
                     , alpha = 0.05)
  
  BIC_res$method <- 'BIC'
  BIC_res$n <- n
  BIC_res$rho <- rho
  
  ## LASSO
  x <- as.matrix(dat[, sprintf("V%02d", 1:20)])
  y <- dat$y
  
  LASSO_redux <- cv.glmnet(x = x
                             , y = y
                             , alpha = 1
                             , type.measure = "mse")

  LASSO_min_res <- harvest_glmnet(
    cv_obj = LASSO_redux,
    param  = param,
    method = "LASSO MIN",
    n      = n,
    rho    = rho,
    s      = "lambda.min"
    )
  
  LASSO_se_res <- harvest_glmnet(
    cv_obj = LASSO_redux,
    param  = param,
    method = "LASSO SE",
    n      = n,
    rho    = rho,
    s      = "lambda.1se"
    )
  
  ## Elastic Net
  EN_redux <- cv.glmnet(x = x
                        , y = y
                        , alpha = 0.5
                        , type.measure = "mse")
  
  EN_min_res <- harvest_glmnet(
    cv_obj = EN_redux,
    param  = param,
    method = "Elastic Net MIN",
    n      = n,
    rho    = rho,
    s      = "lambda.min")
  
   EN_se_res <- harvest_glmnet(
    cv_obj = EN_redux,
    param  = param,
    method = "Elastic Net SE",
    n      = n,
    rho    = rho,
    s      = "lambda.1se")
  
  return(rbind(pval_res, AIC_res, BIC_res, LASSO_min_res, LASSO_se_res, EN_min_res, EN_se_res))
}

test_sim <- simfunc(n = 250, rho = 0.5)
```


## Dealing with many simulation cases

In our simulation we have two main attributes that are being manipulated: sample size and predictor correlation. One way to handle that would be nesting for-loops, but this can get messy fast and it is easy to mix up your indices. One way to cut down the complexity is a profile matrix:

```{r}

profiles <- expand.grid(
  rho = c(0, 0.35, 0.7)
 ,N   = c(250, 500)
)

profiles
```



Now we need a function to iterate through a complete simulation profile.

```{r}

spin <- function(nsim = 500, profile){
  
  stopifnot(nrow(profile) == 1)
  
  res <- list()
  for(iter in 1:nsim) {
    tmp <- simfunc(n = profile[,'N'], rho = profile[,'rho'])
    tmp$iter <- iter
    res[[iter]] <- tmp
  }
  res <- do.call('rbind', res)
  
  return(res)
}
  
test_spin <- spin(nsim = 5, profile = profiles[3,])
```

## The Full Simulation

Defaulting to 10,000 simulations may extend the simulation time, however, estimating a sample size using some reasonable targets may result in a more feasible load (Morris, 2019 - this was uploaded to the week 12 module). For this I will choose to power on the coverage. For this I will use an expected coverage probability of 50% (the worst case in terms of variance). For the SE, I would like to get fairly close, so will use 0.01. This will give reasonable precision in cases of very poor coverage, and much better precision when coverage is closer to 95%. Of course, in critical simulations where precision is critical, higher numbers may be needed and one must give thought to this.

```{r}

# expected coverage
Ep <- 0.5

# required standard error
SEr <- 0.01
nsim <- ceiling(Ep*(1-Ep)/SEr^2)
nsim

# for comparison at 95% coverage and and SE of 0.005 (so we will have better precision than this, where it matters!)
0.95*0.05/(0.005^2)
```



If you are in a hurry, the `future` package can make parallel processing fairly easy:

```{r message = F, warning = F}

n_profiles <- nrow(profiles)

plan(multisession, workers = n_profiles)

set.seed(14)

system.time({
simres <- future_lapply (1:n_profiles, function(i){
  spin(nsim = nsim, profile = profiles[i,])
}
,future.seed=TRUE)
})
simres <- do.call('rbind', simres)
# stop the cluster
plan(sequential)
```

How many 'workers' can you pick? up to the number of cores on your computer but you should leave at least one free for the operating system to continue to function well. You can get a count of available cores:
 8 cores available, should use 7 
```{r}
parallel::detectCores()

```

Compute Coverage probability, by method and simulation setting:

```{r message=F, warning=F}

p_cover <- simres |>
  filter(true_non_zero == 1) |>
  group_by(n, rho, method, variables) |>
  summarise(P_cover = mean(covered)*100)

knitr::kable(
p_cover |> pivot_wider(
  names_from = variables
 ,values_from = P_cover
),
digits = 1)
```

Compute Bias: mean difference between estimated and true coefficient across replicates, treat unselected variables as having an estimate of 0 

Compute true positive rate: proportion of simulations in which each active predictor is retained in the final model
num of v01-v05 retained /5

Compute false positive rate: proportion of sim in which each noise predictor is retained in final model

# References

1. Morris TP, White IR, Crowther MJ. Using simulation studies to evaluate statistical methods. Stat Med. 2019;38(11):2074-2102. doi:10.1002/sim.8086


