---
title: "Simulation"
author: "Rachel Austin"
date: "2026-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(future)
library(future.apply)
library(furrr)
library(hdrm)
library(glmnet)
library(dplyr)
library(openxlsx)
library(tidyr)
library(gt)
```
Potential Cases:
For all 
$\beta_1$ = 0.5/3
$\beta_2$ = 1/3
$\beta_3$ = 1.5/3
$\beta_4$ = 2/3
$\beta_5$ = 2.5/3
$\beta_6 - \beta_20$ = 0

N = 250 and N = 500
  a. $\rho$ = 0
  b. $\rho$ = 0.35
  c. $\rho$ = 0.7
  
Methods:
1. Backwards elimination using $\alpha$ = 0.05
2. Backwards elimination using AIC
3. Backwards elimination using BIC
4. LASSO
  a. $\lambda$.min
  b. $\lambda$.1se
5. Elastic Net


Functions Needed
```{r message = F, warning = F}
# Results extraction function for step function (p-values, AIC and BIC)
harvest <- function(mod_obj      # result model object from the reduction
                   ,param        # simulation target betas
                   ,alpha = 0.05 # crit for coverage and error computation
) {
  
  #extracting coefficients
  estimates <- as.data.frame(summary(mod_obj)$coefficients) 
  #new column called param stores estimate names (previously the rownames)
  estimates <- within(estimates, {
    param <- row.names(estimates)
    })
  #Wald CIs
  CIs <- as.data.frame(confint.default(mod_obj, level = 1-alpha))
  #change name from numeric vals
  names(CIs) <- c('LCL','UCL')
  #new column called param stores estimates names (previously rownames)
  CIs <- within(CIs, {
    param <- row.names(CIs)
  })
  
  #create varname objects
  var_names <- sprintf("V%02d", 1:length(param))
  
  # assemble summaries
  res_dat <- data.frame(
    #varnames of simulated dat
    variables = var_names
    #target betas we assigned in beginning are dropped into true_values
   ,true_values = param
  ) |>
    within({
      # flag whether v01-v20 shows up in estimates, only variables that were not dropped in selection step will show up here
      selected <- ifelse(variables %in% row.names(estimates), 1, 0)
      # flag whether variables v01-v20 are statistically significant 
      signif   <- ifelse(variables %in% row.names(estimates[estimates[,4] < alpha,]), 1, 0)
      # flag the first five var as true non-zero coefs
      true_non_zero <- ifelse(variables %in% sprintf("V%02d", 1:5), 1, 0)
    }) |>
      merge(
    estimates[, c("param", "Estimate")],
    by.x = "variables",
    by.y = "param",
    all.x = TRUE
  ) |>
    merge(CIs
          ,by.x = 'variables'
          ,by.y = 'param'
          ,all.x = T) |>
    within({
      #if the LCL is greater than the target betas assigned and the UCL is less than the target betas assigned we have coverage
      covered <- ifelse(LCL <= true_values & true_values <= UCL, 1, 0)
    })
  
  #rename estimate
  names(res_dat)[names(res_dat) == "Estimate"] <- "estimate"
  #if there is no estimate make it 0
  res_dat$estimate[is.na(res_dat$estimate)] <- 0
    
  #if covered is missing, variable was not selected and variable is a true nonzero effect then covered = 0
  # aka model failed to select a variable that matters 
  res_dat[is.na(res_dat$covered) == T & 
          res_dat$selected == 0 & 
          res_dat$true_non_zero == 1,'covered'] <- 0 
  #if covered is missing, variable was not selected and variable is a true zero effect then covered = 1
  #aka model correctly excluded a null variable
  res_dat[is.na(res_dat$covered) == T & 
          res_dat$selected == 0 & 
          res_dat$true_non_zero == 0,'covered'] <- 1 
  
  return(res_dat)
}

#Results extraction function for glmnet function (LASSO and elastic net)
post_glmnet_lm <- function(cv_obj, dat, s) {
  # allows us to specify whether to pull coefficients for lambda.min model or lambda.1se model
  coefs <- coef(cv_obj, s = s)
  
  # extract var with non-zero coefficients
  selected_vars <- rownames(coefs)[as.numeric(coefs) != 0]
  #exclude intercepts
  selected_vars <- selected_vars[selected_vars != "(Intercept)"]
  
  # if no variables are selected then refit with no variables
  if(length(selected_vars) == 0) {
    return(lm(y ~ 1, data = dat))
  }
  
  # created formula using vars with non-zero coefficients
  form <- as.formula(
    paste("y ~", paste(selected_vars, collapse = " + "))
  )
  
  #refit model with lm
  lm(form, data = dat)
}
   

```



## Sim function: generating data and running analyses
# Initital code provided by Carter Sevick and modified by Rachel Austin
```{r}

simfunc <- function(n, rho) {
  
  # create simulated data
  param <- c(1/6, 1/3, 1/2, 2/3, 5/6, rep(0, 15))
  dat_obj <- gen_data(n = n 
                 ,p = 20
                 ,p1 = 5
                 ,beta = param
                 ,family = 'gaussian'
                 ,corr = 'exchangeable'
                 ,rho = rho
                 ) 
  dat <- data.frame(y=dat_obj$y, dat_obj$X)

  # reduction by p-values, AIC, BIC
  sat_model <- lm(y ~ ., data = dat)

  ## p-values
  pval_redux <- step(sat_model
                    ,direction = 'backward'
                    ,trace = 0
                    ,k = qchisq(1-0.05, 1)
                    )
  pval_res <- harvest(mod_obj = pval_redux  # result model object from the reduction
                     ,param   = param       # simulation target betas
                     ,alpha   = 0.05        # crit for coverage and error computation
  )
  pval_res$method <- 'PVAL'
  pval_res$n      <- n
  pval_res$rho    <- rho
  
  ## AIC
  AIC_redux <- step(sat_model
                    ,direction = 'backward'
                    ,trace = 0
                    ,k = 2
                    )
  AIC_res <- harvest(mod_obj = AIC_redux  # result model object from the reduction
                     ,param   = param       # simulation target betas
                     ,alpha   = 0.05        # crit for coverage and error computation
  )
  AIC_res$method <- 'AIC'
  AIC_res$n      <- n
  AIC_res$rho    <- rho

  ## BIC
  BIC_redux <- step(sat_model
                    , direction = 'backward'
                    , trace = 0
                    , k = log(nrow(dat)))
  
  BIC_res <- harvest(mod_obj = BIC_redux
                     , param = param
                     , alpha = 0.05)
  
  BIC_res$method <- 'BIC'
  BIC_res$n <- n
  BIC_res$rho <- rho
  
  ## LASSO
  x <- as.matrix(dat[, sprintf("V%02d", 1:20)])
  y <- dat$y
  
  LASSO_redux <- cv.glmnet(x = x
                             , y = y
                             , alpha = 1
                             , type.measure = "mse")
  
  ## LASSO min lambda
  LASSO_min_lm <- post_glmnet_lm(LASSO_redux, dat, s = "lambda.min")
  
  LASSO_min_res <- harvest(mod_obj = LASSO_min_lm
                           , param = param
                           , alpha = 0.05
                           )
  
  LASSO_min_res$method <- "LASSO MIN"
  LASSO_min_res$n <- n
  LASSO_min_res$rho <- rho
  
  ## LASSO se lambda
  LASSO_se_lm <- post_glmnet_lm(LASSO_redux, dat, s = "lambda.1se")
  
  LASSO_se_res <- harvest(mod_obj = LASSO_se_lm
                          , param = param
                          , alpha = 0.05
                          )
  
  LASSO_se_res$method <- "LASSO SE"
  LASSO_se_res$n <- n
  LASSO_se_res$rho <- rho
  
  ## Elastic Net
  EN_redux <- cv.glmnet(x = x
                        , y = y
                        , alpha = 0.5
                        , type.measure = "mse")
  
  ## Elastic Net Min
  EN_min_lm <- post_glmnet_lm(EN_redux, dat, s = "lambda.min")
  
  EN_min_res <- harvest(mod_obj = EN_min_lm
                        , param = param 
                        , alpha = 0.05
                        )
  
  EN_min_res$method <- "EN MIN"
  EN_min_res$n <- n
  EN_min_res$rho <- rho
  
  ## Elastic Net SE
  EN_se_lm <- post_glmnet_lm(EN_redux, dat, s = "lambda.1se")
  
  EN_se_res <- harvest(mod_obj = EN_se_lm
                       , param = param
                       , alpha = 0.05)
  
  EN_se_res$method <- "EN SE"
  EN_se_res$n <- n
  EN_se_res$rho <- rho
  
   out <- rbind(pval_res, AIC_res, BIC_res, LASSO_min_res, LASSO_se_res, EN_min_res, EN_se_res)
   out <- out[, c("variables", "true_values","estimate", "true_non_zero", "signif", "selected", "LCL", "UCL", "covered", "method",     "n","rho")]
   return(out)
   }

test_sim <- simfunc(n = 250, rho = 0.5)
```


## Dealing with many simulation cases

```{r}

profiles <- expand.grid(
  rho = c(0, 0.35, 0.7)
 ,N   = c(250, 500)
)

profiles
```


## The Full Simulation

Defaulting to 10,000 simulations may extend the simulation time, however, estimating a sample size using some reasonable targets may result in a more feasible load (Morris, 2019 - this was uploaded to the week 12 module). For this I will choose to power on the coverage. For this I will use an expected coverage probability of 50% (the worst case in terms of variance). For the SE, I would like to get fairly close, so will use 0.01. This will give reasonable precision in cases of very poor coverage, and much better precision when coverage is closer to 95%. Of course, in critical simulations where precision is critical, higher numbers may be needed and one must give thought to this.

```{r}

# expected coverage
Ep <- 0.5

# required standard error
SEr <- 0.01
nsim <- ceiling(Ep*(1-Ep)/SEr^2)
nsim

# for comparison at 95% coverage and and SE of 0.005 (so we will have better precision than this, where it matters!)
0.95*0.05/(0.005^2)
```



```{r}
n_profiles <- nrow(profiles)

plan(multisession, workers = n_profiles)

set.seed(14)

system.time({
  simres_list <- future_lapply(
    1:n_profiles,
    function(i) {
      spin(
        nsim = nsim,
        profile = profiles[i, ]
      )
    },
    future.seed = TRUE
  )
})

simres <- do.call(rbind, simres_list)

plan(sequential)
```

```{r}
write.xlsx(simres, "~/BIOS6624/Project 4/Output/Results.xlsx")
```


Compute true positive rate: proportion of simulations in which each active predictor is retained in the final model
num of v01-v05 retained /5
```{r}
TPR <- simres |>
  filter(true_non_zero == 1) |>
  group_by(n, rho, method) |>
  summarise(
    TPR = mean(selected) * 100,
    .groups = "drop"
  )
```


Compute false positive rate: proportion of sim in which each noise predictor is retained in final model

```{r}
FPR <- simres |>
  filter(true_non_zero == 0) |>
  group_by(n, rho, method) |>
  summarise(
    FPR = mean(selected) * 100,
    .groups = "drop"
  )
```


Compute Bias: mean difference between estimated and true coefficient across replicates, treat unselected variables as having an estimate of 0 
```{r}
Bias <- simres |>
  filter(true_non_zero == 1) |>
  mutate(
    bias = estimate - true_values
  ) |>
  group_by(method, n, rho) |>
  summarise(
    Bias = mean(bias),
    .groups = "drop"
  )
```




Compute Coverage probability, by method and simulation setting:

```{r message=F, warning=F}

Coverage <- simres |>
  filter(true_non_zero == 1) |>
  group_by(method, n, rho) |>
  summarise(
    Coverage = mean(covered),
    .groups = "drop"
  )
```

Compute Type I error: noise variables declared significant
```{r}
Type1 <- simres |>
  filter(true_non_zero == 0) |>
  group_by(method, n, rho) |>
  summarise(
    Type1 = mean(signif),
    .groups = "drop"
  )
```

Compute Type II error: active variables declared not significant
```{r}
Type2 <- simres |>
  filter(true_non_zero == 1) |>
  group_by(method, n, rho) |>
  summarise(
    Type2 = mean(signif == 0),
    .groups = "drop"
  )
```

Compute average model size: average number of selected predictors 
```{r}
ModelSize <- simres |>
  group_by(iter, method, n, rho) |>
  summarise(
    model_size = sum(selected),
    .groups = "drop"
  ) |>
  group_by(method, n, rho) |>
  summarise(
    AvgModelSize = mean(model_size),
    .groups = "drop"
  )
```

Compute prop where all 5 active predictors are selected and no noise predictors are selected
```{r}
CorrectSelection <- simres |>
  group_by(iter, method, n, rho) |>
  summarise(
    all_true = all(selected[true_non_zero == 1] == 1),
    no_false = all(selected[true_non_zero == 0] == 0),
    exact = all_true & no_false,
    .groups = "drop"
  ) |>
  group_by(method, n, rho) |>
  summarise(
    CorrectSelection = mean(exact),
    .groups = "drop"
  )
```


# Tables
```{r}

# Create the simulation scenarios data frame
scenarios <- data.frame(
  Scenario   = 1:6,
  N          = c(250, 250, 250, 500, 500, 500),
  rho        = c(0, 0.35, 0.70, 0, 0.35, 0.70)
)

# Build the gt table
scenarios |>
  gt() |>
  
  # Column labels
  cols_label(
    Scenario = "Scenario",
    N        = md("Sample Size (*N*)"),
    rho      = md("Predictor Correlation (ρ)")
  ) |>
  
  # Title and subtitle
  tab_header(
    title    = "Table 1: Simulation Study Design",
    subtitle = md("Six scenarios crossing sample size and predictor correlation")
  ) |>
  
  # Optional: footnote explaining exchangeable structure
  tab_footnote(
    footnote = "Predictors follow an exchangeable (compound symmetry) correlation structure.",
    locations = cells_column_labels(columns = rho)
  ) |>
  
  # Center numeric columns
  cols_align(align = "center", columns = c(Scenario, rho)) |>
  
  # Format rho as decimals
  fmt_number(columns = rho, decimals = 2)
```


```{r}
FinalResults <- TPR |>
  left_join(FPR, by = c("method", "n", "rho")) |>
  left_join(Bias, by = c("method", "n", "rho")) |>
  left_join(Coverage, by = c("method", "n", "rho")) |>
  left_join(Type1, by = c("method", "n", "rho")) |>
  left_join(Type2, by = c("method", "n", "rho")) |>
  left_join(ModelSize, by = c("method", "n", "rho")) |>
  left_join(CorrectSelection, by = c("method", "n", "rho"))
```

```{r}
FinalResults <- FinalResults |>
  mutate(
    Coverage = Coverage * 100,
    Type1 = Type1 * 100,
    Type2 = Type2 * 100,
    CorrectSelection = CorrectSelection * 100
  )
```


# References

1. Morris TP, White IR, Crowther MJ. Using simulation studies to evaluate statistical methods. Stat Med. 2019;38(11):2074-2102. doi:10.1002/sim.8086


